---
title: "PA1_template"
author: "Raj Purkayastha"
date: "5 May 2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading the data

We will start the analysis by loading the data via read.csv. At this point I will not be modifying the data 

```{r loading, echo=TRUE}
myactivity<-read.csv('activity.csv')

```
##Determining the mean total number of steps taken per day

We will first remove al the NA's present in this data set
```{r removing NAs, echo = TRUE}
mycleanactivity<-myactivity[complete.cases(myactivity),]
```
After this we can use the aggregate function on the new data frame to calculate the total number of steps taken per day

```{r Total number of steps per day, echo = TRUE}
dailysteps<-aggregate(mycleanactivity$steps,FUN=sum,by=list(mycleanactivity$date))
colnames(dailysteps)<-c('Date','Total Steps')

```

The results can be plotted as a histogram using the following code

```{r plotting histogram, echo = TRUE}
hist(dailysteps$`Total Steps`,main='Histogram of Total steps taken each day',xlab='Total Steps',col='blue',xlim=c(0,25000),ylim=c(0,20),breaks=10)

```

The mean and median are calculated as follows

```{r mean and median, echo = TRUE}
dailystepsmean<- mean(dailysteps$`Total Steps`)
dailystepsmedian<-median(dailysteps$`Total Steps`)
```

The mean of daily steps taken is `r dailystepsmean` and the median value is `r dailystepsmedian` .

##Determining the average daily activity pattern

to determine this we will plot a time series plot of the average number of steps taken for each interval across all the days

```{r interval time series, echo = TRUE}
intervalsteps<-aggregate(mycleanactivity$steps,FUN=mean,by=list(mycleanactivity$interval))
colnames(intervalsteps)<-c('Interval','Average Steps')

plot(intervalsteps$Interval,intervalsteps$`Average Steps`,type='l',xlab='Interval Number',ylab='Average number of Steps',lwd=2)

```

The interval which contains the maximum nuber of steps is determined the following way

```{r max steps, echo = TRUE}
maxintervalsteps<-max(intervalsteps$`Average Steps`)
locmaxintervalsteps<-which.max(intervalsteps$`Average Steps`)
maxintervalsteps_val<-intervalsteps$Interval[locmaxintervalsteps]

```

The interval number `r maxintervalsteps_val` has a maximum number of steps which is `r maxintervalsteps`. 

##Imputing missing values

The number of missing values is calculated as follows

```{r missing values, echo = TRUE}
logicals<-is.na(myactivity$steps)
totalna<-sum(logicals)
```

The number of missing values is `r totalna`.

To impute missing values we use the Mice library. Using the mice function we perform the following procedure

```{r imputing}
 library(mice)
tempdata<-mice(myactivity,m=5,maxit=50,meth='pmm',seed=500)
completeactivity<-complete(tempdata,1)
```

m=5 refers to the number of imputed datasets, pmm refers to the method which is predictive Mean matching. 

We then generate a histogram of total number of steps along with mean and median calculations.

```{r complete data set}
ca_dailysteps<-aggregate(completeactivity$steps,FUN=sum,by=list(completeactivity$date))
colnames(ca_dailysteps)<-c('Date','Total Steps')
hist(ca_dailysteps$`Total Steps`,main='Histogram of Total steps taken each day for imputed set',xlab='Total Steps',col='blue',xlim=c(0,25000),ylim=c(0,20),breaks=10)

ca_dailystepsmean<- mean(ca_dailysteps$`Total Steps`)
ca_dailystepsmedian<-median(ca_dailysteps$`Total Steps`)

```
The mean of daily steps taken is `r ca_dailystepsmean` and the median value is `r ca_dailystepsmedian` . These values differ and are lower. This implies that he method used for substitution on the whole, put in lower values than the mean for the NA values. 

##Comparing activity patterns between weekdays and weekends

First we create a new factor variable

```{r weekday}
library(timeDate)
isweekdaycol<-isWeekday(completeactivity$date)
isweekendcol<-isWeekend(completeactivity$date)
completeactivity$wDay<-factor(isweekdaycol,levels=c(FALSE,TRUE),labels=c('Weekend','weekday'))

ca_weekday<-completeactivity[isweekdaycol,]
ca_weekend<-completeactivity[isweekendcol,]

wkday_intervalsteps<-aggregate(ca_weekday$steps,FUN=mean,by=list(ca_weekday$interval))
colnames(wkday_intervalsteps)<-c('Interval','Average Steps')

wkend_intervalsteps<-aggregate(ca_weekend$steps,FUN=mean,by=list(ca_weekend$interval))
colnames(wkend_intervalsteps)<-c('Interval','Average Steps')
```
We then proceed to create a panel plot with average number of steps taken averaged across all weekdays and weekends.

```{r weekday panel plot}


par(mfrow=c(2,1))
plot(wkday_intervalsteps$Interval,wkday_intervalsteps$`Average Steps`,type='l',main='Comparison of Steps on the weekday and weekend',xlab='Interval Number',ylab='Weekday ',lwd=2,col='blue')

plot(wkend_intervalsteps$Interval,wkend_intervalsteps$`Average Steps`,type='l',xlab='Interval Number',ylab='Weekend ',lwd=2,col='red')

```
